---
title: "Proteomics (TMT) Differential Abundance Analysis"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    number_sections: no
    theme: lumen
    toc: yes
    toc_collapsed: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = F,
                      message = F)

library(magrittr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(limma)
library(arrow)
library(readr)
library(QFeatures)
library(dreamerr)
library(yaml,warn.conflicts = FALSE)

```

## Introduction

This Rmd describes our differential abundance analysis (DAA) workflow for proteomics TMT data.

Prior to the DAA the .raw data was processed with Proteome Discoverer.

The starting point for the analysis with this .Rmd are PSM level TMT channel intensities reported by Proteome Discoverer.

## Params

```{r}
INPUT_FILES <- read_yaml('config.yaml')
```

## Read data

```{r,read_data}

samples_df <- readr::read_tsv(INPUT_FILES$samples_file) %>%
  janitor::clean_names() %>%
  dplyr::rename(sample_name = sample) %>%
  mutate(sample_name = stringr::str_replace_all(sample_name,'\\+','_'),
         condition = stringr::str_replace_all(condition,'\\+','_')
         )
contrasts_df <- readr::read_tsv(INPUT_FILES$contrasts_file) %>%
  janitor::clean_names() %>%
  mutate(contrast = stringr::str_replace_all(contrast,'\\+','_'),
         contrast = stringr::str_replace_all(contrast,'vs','-')
         )

# abund
abund_df <- arrow::read_parquet(INPUT_FILES$qprot_abund_file)%>%  
  mutate_if(function(x){class(x)=="integer64"},
              as.integer) %>%
  janitor::clean_names() %>%
  mutate(sample_name = stringr::str_replace_all(sample_name,'-','_'),
         condition = stringr::str_replace_all(condition,'-','_')
         )%>%
  mutate(sample_name = stringr::str_replace_all(sample_name,'\\+','_'),
         condition = stringr::str_replace_all(condition,'\\+','_')
         )

# quan
quan_df <- arrow::read_parquet(INPUT_FILES$qprot_quan_file)%>%   
  mutate_if(function(x){class(x)=="integer64"},
              as.integer) %>%
  janitor::clean_names() %>%
  dplyr::rename(contrast = sample_name) %>%
  mutate(contrast = stringr::str_replace_all(contrast,'\\+','_'),
         contrast = stringr::str_replace_all(contrast,'vs','-')
         )


df <- arrow::read_parquet(INPUT_FILES$qpsms_abund_file) %>%
  bind_cols(select(.$peptide_info_json,-unique_pep)) %>%
  select(-peptide_info_json) %>%
  janitor::clean_names() %>%
  mutate(master_accession = stringr::str_remove(master_accession,';.*'),
         sample_name = stringr::str_replace_all(sample_name,'\\+','_')
         ) %>%
 left_join(
    abund_df %>% distinct(master_accession,
                          gene_name,
                          master_description,
                          contaminant),
      by = 'master_accession'

  ) 
```

### create QFeatures obj

```{r, create_qfeatures_obj}

col_data <- samples_df %>% 
              distinct(sample_name,condition,
                       sample_sm_fid,
                       channel,
                       nominator,
                       denominator,
                       condition,
                       replicate
                       ) %>% 
              tibble::column_to_rownames(var = 'sample_name')


row_id <- 'id'
row_data_cols <- c(row_id,
                 'sequence','quan_info', 
                 'master_accession','master_accessions','gene_name',  
                 'master_description',
                 'contaminant',
                 'percolator_pep','redun',
                 'source',
                 'weight',
                 'sps_mass_matches',
                 'percent_isolation_interference',
                 'average_reporter_sn'
                 #'protein_accessions','parent_protein_accessions','parent_protein_group_count'
                 )

wide_df <- df %>%
  dplyr::select(all_of(row_data_cols),'abund','sample_name')%>%
  tidyr::pivot_wider(names_from = c('sample_name'),
                     values_from = 'abund'
) %>%
  group_by(master_accession) %>%
  mutate(n_psms = n(),
         n_pep = n_distinct(sequence),
         n_uni_pep = n_distinct(sequence[redun == 1]),
         ) %>%
  ungroup



se <- wide_df %>% 
  {readSummarizedExperiment(.,
                            ecol = names(.) %in% rownames(col_data),
                            fnames = row_id
  )}

# add additional exp design columns
colData(se)$sample_name <- colnames(se)
colData(se)$condition <-
  as.vector(set_names(col_data$condition,rownames(col_data))[rownames(colData(se))])
colData(se)$channel <-
  as.vector(set_names(col_data$channel,rownames(col_data))[rownames(colData(se))])
colData(se)$nominator <-
  as.vector(set_names(col_data$nominator,rownames(col_data))[rownames(colData(se))])
colData(se)$denominator <-
  as.vector(set_names(col_data$denominator,rownames(col_data))[rownames(colData(se))])
colData(se)$replicate_nb <-
  as.vector(set_names(col_data$replicate,rownames(col_data))[rownames(colData(se))])

qf <- QFeatures(list(psms = se))
  colData(qf) <- colData(se)

qf %<>% zeroIsNA(i= 'psms')  
  
```

## PSM quantities -\> Protein -\> Quantities

### filter

-   keep protein specific peptides (redun == 1)
-   keep psms assigned *quan_info* status *'passed'* ;
    -   sps_mass_matches \>= 60%
    -   percent_isolation_interference \<= 50%
    -   average_reporter_sn \>= 10
    -   no missing values (i.e. psms with missing reporter ion signals are discarded)

```{r, filter}
# reset qf
kept_assays <- c("psms")
qf <- removeAssay(qf,i = setdiff(names(qf),kept_assays))

qf <- addAssay(qf,SummarizedExperiment::subset(qf[['psms']],
                                      redun == 1 & quan_info == 'passed'),
         name = 'qpsms',
         assayLinks = 
                    AssayLinks(AssayLink('qpsms',
                                         from = 'psms'))
         )
qf <- QFeatures::addAssayLink(qf,
                                  from = 'psms',
                                  to = 'qpsms',
                                  varFrom = 'id',
                                  varTo = 'id'
)

# add q_psm & q_uni_pep to raw data
rowData(qf[['qpsms']]) %<>%
  as.data.frame() %>%
  group_by(master_accession) %>%
  mutate( q_psm = n(),
          q_uni_pep = n_distinct(sequence)
         ) %>%
  ungroup %>%
  DataFrame()


```

### mean center reporter ion quantities per sample

```{r, psm_mean_center}
# reset qf
kept_assays <- c("psms","qpsms")
qf <- removeAssay(qf,i = setdiff(names(qf),kept_assays))

# sample mean center 
qf <- normalize(qf, i = "qpsms",
                   name = "qpsms_norm",
                   method = "div.mean")
```

### protein roll-up

```{r, aggregate }
# reset qf
kept_assays <- c("psms","qpsms",'qpsms_norm')
qf <- removeAssay(qf,i = setdiff(names(qf),kept_assays))


qf <-
    aggregateFeatures(qf,
                      "qpsms_norm",
                       name = "proteins",
                       fcol = "master_accession",
                       fun = base::colMeans,
                       na.rm = TRUE)
```

### Calculate protein log ratios

```{r, ratio_transform_fcns}
ratio_transform <- function(qf,
                            i='eg',
                            name = 'eg_ratio',
                            is_log_scale=F,
                            use_group_design = F,
                            ...
){

  check_value(qf,'class(QFeatures)')
  check_arg(.x1=i,
            .x2 = name,
            .type = "character scalar")
  check_arg(.x1=is_log_scale,
            .x2 = use_group_design,
            .type = "logical scalar")

  # no groups -> sweep mean
  if(!use_group_design){
    if(is_log_scale){
      qf <- QFeatures::sweep(qf,i = i,name = name,MARGIN=1,
                  STATS =log2(rowMeans(2^assay(qf[[i]]), na.rm=T)),
                  FUN = "-"
      )
    }else{
      qf <- QFeatures::sweep(qf,i = i,name = name,MARGIN=1,
                  STATS =rowMeans(assay(qf[[i]]), na.rm=T),
                  FUN = "/"
      )
    }
  }else{
    if(all(c('nominator','denominator') %in% names(colData(qf)))){
      .ratio_transform_grouped_design(qf,i=i,name=name,
                                      is_log_scale = is_log_scale,
                                      ...)
    }else{
      stop("missing colData columns 'nominator','denominator'")
    }
  }
}

.ratio_transform_grouped_design <- function(qf,
                                            i ='eg',
                                            name = 'eg_ratio',
                                            is_log_scale=F,
                                            suppress_messages = T,
                                            suppress_warnings = T
){

  check_value(qf,'class(QFeatures)')
  check_arg(.x1=i,
            .x2 = name,
            .type = "character scalar")
  check_arg(.x1=is_log_scale,
            .x2 = suppress_messages,
            .x3 = suppress_warnings,
            .type = "logical scalar")

  # wide to long to wide
  m <- longFormat(qf[,,i],
                  colvars = c('sample_name','condition',
                              'nominator','denominator')) %>%
    as.data.frame() %>%
    {
      if(is_log_scale) mutate(group_by(.,rowname,nominator),
                              ratio = value - log2(mean(2^value[denominator > 0],
                                                        na.rm=T))
      )
      else mutate(group_by(.,rowname,nominator),
                  ratio = value / mean(value[denominator > 0], na.rm=T)
      )
    } %>%
    `if`(suppress_messages,suppressMessages(.),.) %>%
    `if`(suppress_warnings,suppressWarnings(.),.)

  m %<>%
    tidyr::pivot_wider(id_cols = c('rowname'),
                names_from = 'sample_name',
                values_from = 'ratio') %>%
    ungroup() %>%
    tibble::column_to_rownames() %>%
    as.matrix()

  m <- m[match(rownames(m),rownames(rowData(qf[[i]]))) ,
         match(colnames(m),rownames(colData(qf)))]
  se <- SummarizedExperiment(m,
                             colData =  colData(qf[[i]]),
                             rowData =  rowData(qf[[i]])
  )
  qf <- addAssay(qf,se,
                 name = name,
                 assayLinks =
                   AssayLinks(AssayLink(name,
                                        from = i)
                   ))

  # if se isn't empty
  if(nrow(se) > 0){
    qf <- QFeatures::addAssayLinkOneToOne(qf,
                                  from = i,
                                  to = name)
  }

  return(qf)

}

```

```{r, normalise}
# reset qf
kept_assays <- c("psms","qpsms",'qpsms_norm','proteins')
qf <- removeAssay(qf,i = setdiff(names(qf),kept_assays))

# log transform
qf <- logTransform(qf, i = "proteins",
                      name = "log_proteins")

use_group_design <- 'nominator' %in% names(colData(qf)) &
    'denominator' %in% names(colData(qf))
qf <- ratio_transform(qf,
                        i="log_proteins",
                        name="l2r_proteins",
                        is_log_scale=T,
                        use_group_design = use_group_design
)


```

### Median center protein log ratios

```{r}

# reset qf
kept_assays <- c("psms","qpsms",'qpsms_norm','proteins',
                 'log_proteins',"l2r_proteins"
                 )
qf <- removeAssay(qf,i = setdiff(names(qf),kept_assays))

# sample median center 
qf <- normalize(qf, i = "l2r_proteins",
                   name = "m_l2r_proteins",
                   method = "center.median")

```

### Overview of data processing steps

```{r, plot_qf}
plot(qf)
```

## Differential Abundance Analysis

```{r, daa_fcn}
daa <- function(qf,i='l2r_proteins',
                exp_design_df = NA,
                lm_formula = as.formula('~-1+condition'),
                contrasts = NA
                ){
  
  prots <- qf[[i]]
  
  design <- stats::model.matrix(lm_formula, data=exp_design_df )
  colnames(design) %<>% stringr::str_remove(pattern = "^condition")
  fit <- lmFit(assay(prots), design)
  
  contrast_matrix <- limma::makeContrasts(contrasts=contrasts,levels=design)
  # calculate contrast coefficients
  fit_contrasts <- limma::contrasts.fit(fit,contrast_matrix)
  fit_ebayes <- limma::eBayes(fit_contrasts)
  
  quan_idx_col <- 'master_accession'
  # summarise in long format
  protein_quan_df <-
      fit_ebayes$coefficients %>%
      tibble::as_tibble(rownames =NA) %>%
      tibble::rownames_to_column(var = quan_idx_col) %>%
      tidyr::pivot_longer(!all_of(quan_idx_col),
                          names_to = "contrast",
                          values_to = "l2fc") %>%
      left_join(
        fit_ebayes$p.value %>%
          tibble::as_tibble(rownames =NA) %>%
          tibble::rownames_to_column(var = quan_idx_col) %>%
          tidyr::pivot_longer(!all_of(quan_idx_col),
                              names_to = "contrast",
                              values_to = "pval"),
        by = c(quan_idx_col,'contrast')
    ) %>%
    left_join(
      rowData(qf[[i]]) %>%
      as_tibble() %>%
      select(master_accession,gene_name,master_description,
             contaminant,
             q_psm,q_uni_pep,
             n_psms,n_uni_pep),
      by= 'master_accession'
    ) 
  
  return(protein_quan_df)
  
}
```

```{r, daa}

exp_design_df<- colData(qf) %>% 
                as.data.frame() 

contrasts <- unique(contrasts_df$contrast)

protein_quan_df <- 
    daa(qf,
      i = 'm_l2r_proteins',
      exp_design_df = exp_design_df,
      lm_formula = as.formula('~-1+condition'),
      contrasts = contrasts
    )
 

```

## Results

### Volcanoes

```{r, volcanoes, warning=FALSE}

for(c in contrasts){
  
  p <- protein_quan_df %>%
    dplyr::filter(contrast %in% !!c) %>%
    ggplot2::ggplot(ggplot2::aes(x= l2fc, y = -log10(pval))) +
    ggplot2::geom_point()  +
    #ggplot2::geom_hline(yintercept = 2, linetype = 'dashed')+
    #ggplot2::ggtitle(c)+
    ggplot2::theme_bw()
  
  plot(p)
}

```

## Export

```{r, export, eval =F}
readr::write_tsv(protein_quan_df,file = paste0(tempdir(),'/protein_daa.tsv'))
```

## Check

The data was originally processed by an in-house pipeline (*org_pipline*). Here we confirm that this .Rmd generates identical results.

### protein stats

```{r}
rowData(qf[['proteins']]) %>%
          as_tibble() %>%
  left_join(
    abund_df %>%
      dplyr::rename(n_psms = n_ps_ms) %>%
      distinct(master_accession,
               q_psm,
               n_psms,
               n_pep,
               n_uni_pep,
               q_uni_pep),
    suffix = c('_qf','_org_pipeline'),
    by = 'master_accession'
  ) %>%
  #mutate(check_q_psm =  ) %>%
  summarise(all(q_psm_org_pipeline == q_psm_qf),
            all(n_psms_org_pipeline == n_psms_qf),
            all(n_pep_org_pipeline == n_pep_qf),
            all(n_uni_pep_org_pipeline == n_uni_pep_qf),
            all(q_uni_pep_org_pipeline == q_uni_pep_qf)
            ) %>%
  all() %>%
  testthat::expect_true()
  
```

### abund

```{r, check_abund}

longFormat(qf[,,'proteins'],
                        colvars = c('condition'),
                        rowvars = 'master_accession'
                        ) %>%
  as.data.frame() %>%
  dplyr::rename(sample_name = colname,
         abund = value) %>%
  select('sample_name','master_accession', 'abund') %>%
  left_join(
    abund_df,
    suffix = c('_qf','_org_pipeline'),
    by = c('sample_name','master_accession')
  ) %>%
  select(master_accession,sample_name,abund_org_pipeline,abund_qf) %>%
  # mutate(abund_org = signif(abund_org,4),
  #        abund_qf = signif(abund_qf,4)
  #        ) %>%
  # mutate(check = abund_org == abund_qf) %>%
  sample_n(1000) %>%
  ggplot(aes(x = log(abund_org_pipeline), y= log(abund_qf)))+
  geom_point()
```

### m_l2r

```{r, check_m_l2r}
longFormat(qf[,,'m_l2r_proteins'],
                        colvars = c('condition'),
                        rowvars = 'master_accession'
                        ) %>%
  as.data.frame() %>%
  dplyr::rename(sample_name = colname,
         m_l2r = value) %>%
  select('sample_name','master_accession', 'm_l2r') %>%
  left_join(
    abund_df,suffix = c('_qf','_org_pipeline'),
    by = c('sample_name','master_accession')
  ) %>%
  select(master_accession,sample_name,m_l2r_org_pipeline,m_l2r_qf) %>%
  # mutate(abund_org = signif(abund_org,4),
  #        abund_qf = signif(abund_qf,4)
  #        ) %>%
  # mutate(check = abund_org == abund_qf) %>%
  sample_n(1000) %>%
  ggplot(aes(x = m_l2r_org_pipeline, y= m_l2r_qf))+
  geom_point()

# check qProts
testthat::expect_equal(nrow(qf[['m_l2r_proteins']]),
  filter(abund_df,!is.na(m_l2r))$master_accession %>% n_distinct())

```

### l2fc, pval

```{r, check daa}
comp_df <- protein_quan_df %>%
  select(contrast,master_accession,pval,l2fc) %>%
  left_join(
    quan_df %>%
      select(contrast,master_accession,pval,l2fc),
    suffix = c('_qf','_org_pipeline'),
    by = c('contrast','master_accession')
  ) 

comp_df %>%
  #sample_n(5000) %>%
  ggplot(aes(x = l2fc_org_pipeline, y= l2fc_qf))+
  geom_point()

comp_df %>%
  sample_n(5000) %>%
  ggplot(aes(x = pval_org_pipeline, y= pval_qf))+
  geom_point()

# check qProts
testthat::expect_equal(filter(protein_quan_df,!is.na(pval))$master_accession %>% n_distinct(),
  filter(quan_df,!is.na(pval))$master_accession %>% n_distinct())

```

## Session information

```{r, eval=F}
sessionInfo()
```
